Calculating life expectancy and exploring missingness

### 2017-01-20 Code/1_Explore_Data.Rmd
change var format

```{r read_raw_data}
dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
# names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
```

```{r recode_variables}
dat <- dat %>%
  mutate(State_factor = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age_factor = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age_minbin = 0*(Age_factor == "<1 year") + 1*(Age_factor == "1-4 years") + 5*(Age_factor == "5-9 years") +
                10*(Age_factor == "10-14 years") + 15*(Age_factor == "15-19 years") + 20*(Age_factor == "20-24 years") + 
                25*(Age_factor == "25-29 years") + 30*(Age_factor == "30-34 years") + 35*(Age_factor == "35-39 years") +
                40*(Age_factor == "40-44 years") + 45*(Age_factor == "45-49 years") + 50*(Age_factor == "50-54 years") + 
                55*(Age_factor == "55-59 years") + 60*(Age_factor == "60-64 years") + 65*(Age_factor == "65-69 years") +   
                70*(Age_factor == "70-74 years") + 75*(Age_factor == "75-79 years") + 80*(Age_factor == "80-84 years") + 
                85*(Age_factor == "85+ years"),
         Age_bin_size = 1*(Age_factor == "<1 year") + 4*(Age_factor == "1-4 years") + 5*(Age_factor != "<1 year" & Age_factor != "1-4 years"),
         Sex_factor = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race_factor = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD_factor = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                        "Non-communicable", "Injuries", "All other causes")), 
         Year_factor = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         Year_actual = as.numeric(as.character(Year_factor)),
         RaceSex = interaction(Race_factor, Sex_factor, sep = " "),
         StateYearSex = interaction(State_factor, Year_factor, Sex_factor),
         StateYearRaceSex = interaction(State_factor, Year_factor, Race_factor, Sex_factor))

@@

```{r remove_unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dat$Age_factor)

#Count of death is always equal to 0 for unknown age groups
table(dat$Age_factor, dat$Count==0)

#remove these data from the data we analyze:
dat.clean <- dat[dat$Age_factor != "Unknown", ]
```

@@

```{r recode_pops_of_zero}
table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0.

table(dat.clean$Count[dat.clean$Population == 0] == 0, useNA = "always")
#Of these, all have Counts of 0, except for 26 with counts of NA

#replace the NA Counts with 0 for the 26 cells with Populations of 0.
dat.clean$Count[is.na(dat.clean$Count) == T & dat.clean$Population == 0] <- 0
summary(dat.clean$Count[dat.clean$Population == 0])

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?
```

```{r list_states_0_le20_popcounts}
#States with at least one strata of 0
na.pops <- dat.clean %>% 
  select(State_factor, Population) %>% 
  filter(Population == 0)

unique(na.pops$State_factor)

pops.le.20 <- dat.clean %>% select(State_factor, Population) %>% filter(Population <= 20)
unique(pops.le.20$State_factor)
```

@@

```{r fill_in_missing_counts}
dat.clean <- dat.clean %>% mutate(Count_md1 = ifelse(is.na(Count)==T, pmin(1, Population), Count),
                                  Count_md5 = ifelse(is.na(Count)==T, pmin(5, Population), Count),
                                  Count_md9 = ifelse(is.na(Count)==T, pmin(9, Population), Count)
                                  )

#check recoding
dat.clean$Count_grouped <- cut(dat.clean$Count, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count1_grouped <- cut(dat.clean$Count_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count5_grouped <- cut(dat.clean$Count_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$Count9_grouped <- cut(dat.clean$Count_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))

table(dat.clean$Count_grouped, dat.clean$Count1_grouped, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count5_grouped, useNA = "always")
table(dat.clean$Count_grouped, dat.clean$Count9_grouped, useNA = "always")

dat.clean <- dat.clean[ , !names(dat.clean) %in% c("Count_grouped", "Count1_grouped", "Count5_grouped", "Count9_grouped")]
```

@@

dat.clean <- dat.clean %>% group_by(State_factor, RaceSex, Year_actual, Age_minbin) %>%
                           arrange(State_factor, RaceSex, Year_actual, Age_minbin) %>%
                           mutate(total_deaths1 = sum(Count_md1),
                                  total_deaths5 = sum(Count_md5),
                                  total_deaths9 = sum(Count_md9),
                                  cod_prop_death1 = Count_md1/total_deaths1, 
                                  cod_prop_death5 = Count_md5/total_deaths5,
                                  cod_prop_death9 = Count_md9/total_deaths9)

@@

```{r aggregate_over_COD}
dat.aggregated <- dat.clean %>% group_by(State_factor, Year_factor, RaceSex, Age_factor) %>%
                                arrange(State_factor, RaceSex, Year_factor, Age_factor) %>%
                                summarise(Population = first(Population), 
                                          Race_factor = first(Race_factor), 
                                          Sex_factor = first(Sex_factor), 
                                          StateYearRaceSex = first(StateYearRaceSex),
                                          Age_bin_size = first(Age_bin_size),
                                          Age_minbin = first(Age_minbin),
                                          Year_actual = first(Year_actual),
                                          total_deaths1 = first(total_deaths1),
                                          total_deaths5 = first(total_deaths5),
                                          total_deaths9 = first(total_deaths9),
                                          cod_prop_death1 = first(cod_prop_death1),
                                          cod_prop_death5 = first(cod_prop_death5),
                                          cod_prop_death9 = first(cod_prop_death9))

#return "stratum ids" with any age-specific population counts of 0.
non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))
```

@@

```{r create_var_pop_across_age}
dat3 <- dat.aggregated %>% ungroup() %>% 
                           group_by(StateYearRaceSex) %>%
                           summarise(Pop_across_age = sum(Population))

dat.aggregated <- merge(dat.aggregated, dat3, by = "StateYearRaceSex")

@@

dat.aggregated <- dat.aggregated %>% arrange(State_factor, Year_actual, Race_factor, Sex_factor, Age_minbin)
```

```{r write_data_to_csvs}
#save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.csv", col_names = T)
write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.csv", col_names = T)

```

>>>>

```{r read_raw_data}
dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
names(dat) <- c("state.n", "age.n", "sex.n", "race.n", "COD.n", "year.0", "crude.rate", "deaths", "population") 
# names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
```

```{r recode_variables}
dat <- dat %>%
  mutate(state = factor(state.n, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         age = factor(age.n, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         age_minbin = 0*(age == "<1 year") + 1*(age == "1-4 years") + 5*(age == "5-9 years") +
                10*(age == "10-14 years") + 15*(age == "15-19 years") + 20*(age == "20-24 years") + 
                25*(age == "25-29 years") + 30*(age == "30-34 years") + 35*(age == "35-39 years") +
                40*(age == "40-44 years") + 45*(age == "45-49 years") + 50*(age == "50-54 years") + 
                55*(age == "55-59 years") + 60*(age == "60-64 years") + 65*(age == "65-69 years") +   
                70*(age == "70-74 years") + 75*(age == "75-79 years") + 80*(age == "80-84 years") + 
                85*(age == "85+ years"),
         age_bin_size = 1*(age == "<1 year") + 4*(age == "1-4 years") + 5*(age != "<1 year" & age != "1-4 years"),
         sex = factor(sex.n, levels = c(0, 1), labels = c("Male", "Female")),
         race = factor(race.n, levels = c(0, 1), labels = c("White", "Black")),
         COD = factor(COD.n, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                        "Non-communicable", "Injuries", "All other causes")), 
         year = factor(year.0, levels = c(0:44), labels = as.character(1969:2013)),
         year.n = as.numeric(as.character(year)),
         race.sex = interaction(race, sex, sep = " "),
         state.year.sex = interaction(state, year, sex),
         state.year.race.sex = interaction(state, year, race, sex))

@@

```{r remove_unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dat$age)

#Count of death is always equal to 0 for unknown age groups
table(dat$age, dat$deaths==0)

#remove these data from the data we analyze:
dat.clean <- dat[dat$age != "Unknown", ]
```

@@

```{r recode_pops_of_zero}
table(dat.clean$population == 0, useNA = "always")
#654 cells have Populations of 0.

table(dat.clean$deaths[dat.clean$population == 0] == 0, useNA = "always")
#Of these, all have Counts of 0, except for 26 with counts of NA

#replace the NA Counts with 0 for the 26 cells with Populations of 0.
dat.clean$deaths[is.na(dat.clean$deaths) == T & dat.clean$population == 0] <- 0
summary(dat.clean$deaths[dat.clean$population == 0])

summary(dat.clean$crude.rate[dat.clean$population == 0])
#whenever the population is 0, the crude.rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?
```

```{r list_states_0_le20_popcounts}
#States with at least one strata of 0
na.pops <- dat.clean %>% 
  select(state, population) %>% 
  filter(population == 0)

unique(na.pops$state)

pops.le.20 <- dat.clean %>% select(state, population) %>% filter(population <= 20)
unique(pops.le.20$state)
```

@@

```{r fill_in_missing_counts}
dat.clean <- dat.clean %>% mutate(deaths_md1 = ifelse(is.na(deaths)==T, pmin(1, population), deaths),
                                  deaths_md5 = ifelse(is.na(deaths)==T, pmin(5, population), deaths),
                                  deaths_md9 = ifelse(is.na(deaths)==T, pmin(9, population), deaths)
                                  )

#check recoding
dat.clean$deaths_grouped <- cut(dat.clean$deaths, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$deaths1_grouped <- cut(dat.clean$deaths_md1, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$deaths5_grouped <- cut(dat.clean$deaths_md5, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))
dat.clean$deaths9_grouped <- cut(dat.clean$deaths_md9, c(-1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 22040))

table(dat.clean$deaths_grouped, dat.clean$deaths1_grouped, useNA = "always")
table(dat.clean$deaths_grouped, dat.clean$deaths5_grouped, useNA = "always")
table(dat.clean$deaths_grouped, dat.clean$deaths9_grouped, useNA = "always")

dat.clean <- dat.clean[ , !names(dat.clean) %in% c("deaths_grouped", "deaths1_grouped", "deaths5_grouped", "deaths9_grouped")]
```

@@

dat.clean <- dat.clean %>% group_by(state, race.sex, year.n, age_minbin) %>%
                           arrange(state, race.sex, year.n, age_minbin) %>%
                           mutate(total_deaths1 = sum(deaths_md1),
                                  total_deaths5 = sum(deaths_md5),
                                  total_deaths9 = sum(deaths_md9),
                                  cod_prop_death1 = deaths_md1/total_deaths1, 
                                  cod_prop_death5 = deaths_md5/total_deaths5,
                                  cod_prop_death9 = deaths_md9/total_deaths9)

@@

```{r aggregate_over_COD}
dat.aggregated <- dat.clean %>% group_by(state, year, race.sex, age) %>%
                                arrange(state, race.sex, year, age) %>%
                                summarise(population = first(population), 
                                          race = first(race), 
                                          sex = first(sex), 
                                          state.year.race.sex = first(state.year.race.sex),
                                          age_bin_size = first(age_bin_size),
                                          age_minbin = first(age_minbin),
                                          year.n = first(year.n),
                                          total_deaths1 = first(total_deaths1),
                                          total_deaths5 = first(total_deaths5),
                                          total_deaths9 = first(total_deaths9),
                                          cod_prop_death1 = first(cod_prop_death1),
                                          cod_prop_death5 = first(cod_prop_death5),
                                          cod_prop_death9 = first(cod_prop_death9))

#return "stratum ids" with any age-specific population counts of 0.
non.zero.stratums <- setdiff(levels(dat.aggregated$state.year.race.sex), 
                             unique(dat.aggregated$state.year.race.sex[dat.aggregated$population == 0]))
```

@@

```{r create_var_pop_across_age}
dat3 <- dat.aggregated %>% ungroup() %>% 
                           group_by(state.year.race.sex) %>%
                           summarise(pop_across_age = sum(population))

dat.aggregated <- merge(dat.aggregated, dat3, by = "state.year.race.sex")

@@

dat.aggregated <- dat.aggregated %>% arrange(state, year.n, race, sex, age_minbin)
```

```{r write_data_to_csvs}
#save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)
```
